---
title: "Alpha diversity"
output: html_notebook
---

# Alpha Diversity


## Load packages needed
```{r}
library("ggplot2")
library("multcompView")
library("patchwork")
library("phyloseq")
library("plotrix")
library("tidyverse")
library("vegan")
```


## Load the files needed
```{r}
ps <- readRDS("examplePhyloseq.rds") #import the .rds phyloseq object made in the intro to phyloseq notebook
ps #confirm it worked
lowps<-subset_samples(ps,Region=="Lower Keys")
lowps
midps<-subset_samples(ps,Region=="Middle Keys")
midps
upps<-subset_samples(ps,Region=="Upper Keys")
upps
```


## Calculate and Save Alpha Diversity Metrics
```{r}
myalphamets<-function(p){
    asvnew<-as.matrix((data.frame(otu_table(p))))#convert otu table to matrix
  samdf<-data.frame(sample_data(p))#convert sample data to a dataframe
  rich1 <- specnumber(asvnew)#calculates richness from css
  length(rich1)#check the length of the richness calculated
  shannon1 <- vegan::diversity(asvnew,"shannon")#calculates the shannon diversity using the function in vegan
  length(shannon1)#check the length of the shannon diversity calculated
  invsimp1 <- vegan::diversity(asvnew, "inv")#calculates the inverse simpson diversity using the function in vegan
  length(invsimp1)#check the length of the inverse simpson diversity calculated
  alp <- data.frame()#make empty dataframe to put the calculated alpha diversity into
  alp <- cbind(rich1, shannon1, invsimp1,samdf)#put the calculated alpha diversity into the dataframe
  alp <- rownames_to_column(alp, var = "id")
  alp <- data.frame(alp)#back to dataframe from matrix
  alp#confirm it works
}
alpha_coral<-myalphamets(ps)
alpha_coral
lower_alpha<-myalphamets(lowps)
lower_alpha
middle_alpha<-myalphamets(midps)
middle_alpha
upper_alpha<-myalphamets(upps)
upper_alpha
```


##Plotting Alpha Diversity Metrics


### Plot Richness
```{r}
  ric <- alpha_coral %>% ggplot(aes(Region,rich1)) + geom_boxplot() + xlab("Region") +  theme(text = element_text(size=12))+ylab("Species Richness") + ggtitle("Richness") + theme(axis.text.x = element_text(angle = 90,size=12)) +  theme(legend.text=element_text(size=12), legend.position = "right",text = element_text(size=12),strip.text= element_text(size=12)) + theme(axis.text.x = element_text(size=12))
  ric
```


### Plot Shannon Diversity
```{r}
shandiv <- alpha_coral %>% ggplot(aes(Region,shannon1)) + geom_boxplot() + ggtitle("Shannon Diversity") + ylab("Shannon")+xlab("Region")+ylab("Shannon Diversity") + theme(text = element_text(size=10)) + theme(axis.text.x = element_text(angle = 90)) + theme(legend.text=element_text(size=12),  legend.position = "right",text = element_text(size=12),strip.text= element_text(size=12)) + theme(axis.text.x = element_text(size=12))
  shandiv
```


### Plot Inverse Simpson Diversity
```{r}
invsimp <- alpha_coral %>% ggplot(aes(Region, invsimp1)) + geom_boxplot() + ylab("Inverse Simpson Diversity") + xlab("Region") + theme(legend.position="none") + theme(text = element_text(size=12)) + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Inverse Simpson Diversity")  +theme(legend.text=element_text(size=12), legend.position = "right",text = element_text(size=12),strip.text= element_text(size=12)) + theme(axis.text.x = element_text(size=12))
  invsimp
```

#put the graphs together, i.e. make it pretty
```{r}
ric + shandiv + invsimp + plot_layout(guides="collect")
```


## average, range, standard deviation, and standard error of the alpha diversity metrics
```{r}
avgalpha <- function(alpha){
    avg<-data.frame(matrix(nrow = 4,ncol = 6))
    avg[1, ]<-c("alpha diversity metric","Mean","Median","range","standard deviation","standard error")
    avg[ ,1]<-c("alpha diversity metric","richness","shannon","inv simp")
    avg[2,2]<-mean(alpha$rich1)
    avg[2,3]<-median(alpha$rich1)
    range<-data.frame(range(alpha$rich1))
    range<-as.character(range)
    avg[2,4]<- as.character(interaction(range,sep="_"))
    avg[2,5]<-sd(alpha$rich1)
    avg[2,6]<-std.error(alpha$rich1)
    
    avg[3,2]<-mean(alpha$shannon1)
    avg[3,3]<-median(alpha$shannon1)
    range<-data.frame(range(alpha$shannon1))
    range<-as.character(range)
    avg[3,4]<- as.character(interaction(range,sep="_"))
    avg[3,5]<-sd(alpha$shannon1)
    avg[3,6]<-std.error(alpha$shannon1)
    
    avg[4,2]<-mean(alpha$invsimp1)
    avg[4,3]<-median(alpha$invsimp1)
    range<-data.frame(range(alpha$invsimp1))
    range<-as.character(range)
    avg[4,4]<- as.character(interaction(range,sep="_"))
    avg[4,5]<-sd(alpha$invsimp1)
    avg[4,6]<-std.error(alpha$invsimp1)
    avg
}
lowavgalp<-avgalpha(lower_alpha)
lowavgalp
midavgalp<-avgalpha(middle_alpha)
midavgalp
upavgalp<-avgalpha(upper_alpha)
upavgalp
```


## check normality
```{r}
mod<-aov((rich1)~Region,data=alpha_coral) #no need for a transformation
shapiro.test(mod$residuals)
summary(mod)
tuk.rich<-TukeyHSD(mod)
tuk.rich
multcompLetters(tuk.rich$Region[, "p adj"])$Letters

mod<-aov((shannon1)~Region,data=alpha_coral) #no need for a transformation
shapiro.test(mod$residuals)
summary(mod)
tuk.rich<-TukeyHSD(mod)
tuk.rich
multcompLetters(tuk.rich$Region[, "p adj"])$Letters

mod<-aov((invsimp1)~Region,data=alpha_coral) #no need for a transformation
shapiro.test(mod$residuals)
summary(mod)
tuk.rich<-TukeyHSD(mod)
tuk.rich
multcompLetters(tuk.rich$Region[, "p adj"])$Letters
```


## kruskal wallis
```{r}
alpha_coral$Region<-as.factor(alpha_coral$Region)
al2<-data.frame()
a1<-(kruskal.test(rich1~Region,data=alpha_coral))
a1list<-c(a1$data.name,a1$parameter,a1$statistic,a1$p.value)
a2<-(kruskal.test(shannon1~Region,data=alpha_coral))
a2list<-c(a2$data.name,a2$parameter,a2$statistic,a2$p.value)
a3<-(kruskal.test(invsimp1~Region,data=alpha_coral))
a3list<-c(a3$data.name,a3$parameter,a3$statistic,a3$p.value)
ak<-data.frame(a1list,a2list,a3list)
colnames(ak)<-ak[1,]
rownames(ak)<-c("data.name","parameter","statistic","p.value")
ak
```


## multiple kruskal wallis
#### richness pairwise kruskal wallis
```{r}
sampledf<-data.frame(sample_data(ps))
sampledf$Region<-as.factor(sampledf$Region)
region.list<-levels(sampledf$Region)
region.list
aovresults<-c()
pvals<-c()
locations<-c()
for (i in 1:length(region.list)){
  region.leftout<-region.list[i]
  alph_sub<-subset(alpha_coral,Region!=region.leftout) #instead of subset_samples use a subset() for the dataframe you are working with
  #d.sub<-phyloseq::distance(ps_sub, "bray") # won't need this for kruskal
  #df.new<-subset(sampledf,subset=Region!=region.leftout) # won't need this either
  PERM<-kruskal.test(rich1~Region,alph_sub) #change to kruskal
  print(PERM)
  pvals<-c(pvals,PERM$'p.value') #this will change. run a kruskal and save the output to do str(output) to see what it call's the pvalues that it outputs.
  #aovresults<-rbind(aovresults,PERM$aov.tab) 
  loc.sub<-region.list[-i]
  locations<-rbind(locations,c(loc.sub))
}
aovresults
combinedtest<-cbind(locations,pvals)
mult_perm<-data.frame(combinedtest)
mult_perm
mult_perm$pvaladj<-p.adjust(mult_perm$pvals,method="bonferroni")
mult_perm
```

#### shannon diversity pairwise kruskal wallis
```{r}
sampledf<-data.frame(sample_data(ps))
sampledf$Region<-as.factor(sampledf$Region)
region.list<-levels(sampledf$Region)
region.list
aovresults<-c()
pvals<-c()
locations<-c()
for (i in 1:length(region.list)){
  region.leftout<-region.list[i]
  alph_sub<-subset(alpha_coral,Region!=region.leftout) #instead of subset_samples use a subset() for the dataframe you are working with
  #d.sub<-phyloseq::distance(ps_sub, "bray") # won't need this for kruskal
  #df.new<-subset(sampledf,subset=Region!=region.leftout) # won't need this either
  PERM<-kruskal.test(shannon1~Region,alph_sub) #change to kruskal
  print(PERM)
  pvals<-c(pvals,PERM$'p.value') #this will change. run a kruskal and save the output to do str(output) to see what it call's the pvalues that it outputs.
  #aovresults<-rbind(aovresults,PERM$aov.tab) 
  loc.sub<-region.list[-i]
  locations<-rbind(locations,c(loc.sub))
}
aovresults
combinedtest<-cbind(locations,pvals)
mult_perm<-data.frame(combinedtest)
mult_perm
mult_perm$pvaladj<-p.adjust(mult_perm$pvals,method="bonferroni")
mult_perm
```

#### inverse simpson pairwise kruskal wallis
```{r}
sampledf<-data.frame(sample_data(ps))
sampledf$Region<-as.factor(sampledf$Region)
region.list<-levels(sampledf$Region)
region.list
aovresults<-c()
pvals<-c()
locations<-c()
for (i in 1:length(region.list)){
  region.leftout<-region.list[i]
  alph_sub<-subset(alpha_coral,Region!=region.leftout) #instead of subset_samples use a subset() for the dataframe you are working with
  #d.sub<-phyloseq::distance(ps_sub, "bray") # won't need this for kruskal
  #df.new<-subset(sampledf,subset=Region!=region.leftout) # won't need this either
  PERM<-kruskal.test(invsimp1~Region,alph_sub) #change to kruskal
  print(PERM)
  pvals<-c(pvals,PERM$'p.value') #this will change. run a kruskal and save the output to do str(output) to see what it call's the pvalues that it outputs.
  #aovresults<-rbind(aovresults,PERM$aov.tab) 
  loc.sub<-region.list[-i]
  locations<-rbind(locations,c(loc.sub))
}
aovresults
combinedtest<-cbind(locations,pvals)
mult_perm<-data.frame(combinedtest)
mult_perm
mult_perm$pvaladj<-p.adjust(mult_perm$pvals,method="bonferroni")
mult_perm
```

