---
title: "Alpha diversity"
output: html_notebook
---

# Alpha Diversity

Alpha diversity describes the microbiome diversity *within* a sample/group. Common measures of alpha diversity are:

- Richness
- Evenness
- Shannon Diversity Index
- Inverse Simpson Diversity Index

See the vegan::diversity-vegan R vignette for details on metric calculation.

## Load packages needed
```{r}
library("ggplot2") #for plotting
library("multcompView") #for Tukey post-hoc test
library("patchwork") #for plotting
library("phyloseq") #for microbial data
library("tidyverse") #for data filtering
library("vegan") #for stats
```


## Load the files needed
```{r}
ps <- readRDS("examplePhyloseq.rds") #import the .rds phyloseq object made in the intro to phyloseq notebook
ps #confirm it worked
```


## Calculate and Save Alpha Diversity Metrics
```{r}
#get data into correct format for calculation
asvtab<-as.matrix((data.frame(otu_table(ps))))#convert otu table to matrix
samdf<-data.frame(sample_data(ps))#convert sample data to a dataframe

#richness
rich <- specnumber(asvtab)
length(rich)#check the length, should be the same as the number of samples

#Shannon diversity
shan.div <- vegan::diversity(asvtab,"shannon")

#Inverse Simpson
inv.simp <- vegan::diversity(asvtab, "inv")

#combine them all into one dataframe with the sample data
alp <- cbind(rich, shan.div, inv.simp,samdf)
alp <- rownames_to_column(alp, var = "id") #add the rownames as a column
alphamets <- data.frame(alp)#back to dataframe from matrix
alphamets

#use the function wrapper from the AlphaDiv_wrapperfuncs.R script to save yourself some time (and lines of code)
source("AlphaDiv_wrapperfuncs.R")

alpha_coral<-myalphamets(ps) #run the function on the full dataset and save the output
alpha_coral #check it worked
```
## Statistics

### average and standard error of the alpha diversity metrics
```{r}
#If you just want the mean and the standard error of one metric at a time:
aggregate(rich1~Region,FUN=mean,data=alpha_coral) #mean
aggregate(rich1~Region,FUN=function(x) sd(x)/sqrt(length(x)),data=alpha_coral) #se

#now use the wrapper function to get everything at once
alpha.summary(ps,sample_data(ps)$Region) #must put in phyloseq object and then the sampledata variable that you are interested in

```

### Does alpha diversity vary by region?

Use anovas (analysis of variance model) to test for significant differences among groups.
```{r}

####### richness: ##########

#the anova model
mod<-aov((rich1)~Region,data=alpha_coral)

#test for normality using shapiro-wilks
shapiro.test(mod$residuals) #if p<0.05, then the data is not NORMAL
#it wasn't normal
mod<-aov(log(rich1)~Region,data=alpha_coral) #try a log transformation
shapiro.test(mod$residuals) #now we're normal.

#get anova results
summary(mod) #region is significantly different

#do the multiple comparison tests (tukey)
tuk.rich<-TukeyHSD(mod)
tuk.rich #the lower keys are sig different from middle and upper
rich.letters<-multcompLetters(tuk.rich$Region[, "p adj"])$Letters #save letters for plotting

###### shannon: #######
mod<-aov((shannon1)~Region,data=alpha_coral) 
shapiro.test(mod$residuals) #normal
summary(mod) #region is significant
tuk.shan<-TukeyHSD(mod)
tuk.shan #upper sig dif from lower
shan.letters<-multcompLetters(tuk.shan$Region[, "p adj"])$Letters

####### inverse simpson: #######
mod<-aov((invsimp1)~Region,data=alpha_coral)
shapiro.test(mod$residuals) #Not normal
mod<-aov(log(invsimp1)~Region,data=alpha_coral) #try with log transform
shapiro.test(mod$residuals) #Now we're good
summary(mod) #region is sig dif
tuk.simp<-TukeyHSD(mod)
tuk.simp #lower sig dif from middle and upper
simp.letters<-multcompLetters(tuk.simp$Region[, "p adj"])$Letters
```

## Plotting Alpha Diversity Metrics

### Plot Richness
```{r}
#make a box plot with region on the x-axis, and species richness on the y-axis, saved in a variable so that it can be saved later or used with patchwork

ric <- alpha_coral %>% ggplot(aes(Region,rich1)) + 
  geom_boxplot() + 
  xlab("") +  
  theme(text = element_text(size=12))+
  ylab("Species Richness") + 
  #ggtitle("Richness") + 
  theme(axis.text.x = element_text(angle = 90))+
  annotate(geom="text",x=c(1,2,3),y=c(70,70,70),label=rich.letters[order(names(rich.letters))],color="red")
#you will need to adjust the annotate line individually, as I took the easy way out when coding this instead of making it work with anything
ric #print the plot
```

### Plot Shannon Diversity
```{r}
#make a box plot with region on the x-axis, and shannon diversity on the y-axis, saved in a variable so that it can be saved later with ggsave
shandiv <- alpha_coral %>% ggplot(aes(Region,shannon1)) + 
  geom_boxplot() + 
  #ggtitle("Shannon Diversity") + 
  xlab("")+
  ylab("Shannon Diversity") + 
  theme(text = element_text(size=12)) + 
  theme(axis.text.x = element_text(angle = 90))+
  annotate(geom="text",x=c(1,2,3),y=c(0.75,0.75,0.75),label=shan.letters[order(names(shan.letters))],color="red")

shandiv #print the plot
```


### Plot Inverse Simpson Diversity
```{r}
#make a box plot with region on the x-axis, and species richness on the y-axis, saved in a variable so that it can be saved later with ggsave

invsimp <- alpha_coral %>% ggplot(aes(Region, invsimp1)) + 
  geom_boxplot() + 
  ylab("Inverse Simpson Diversity") + 
  xlab("") + 
  theme(text = element_text(size=12)) + 
  theme(axis.text.x = element_text(angle = 90))+
    annotate(geom="text",x=c(1,2,3),y=c(1.7,1.7,1.7),label=simp.letters[order(names(simp.letters))],color="red")

invsimp #print the plot
```

#### put the graphs together into one multi-panel figure using Patchwork
```{r}
ric + shandiv + invsimp 
```
## Back to stats real quick.

If your data are very not normal and the usual transformations are not working, then you will need to resort to using a kruskal wallis test, the non-parametric version of an ANOVA.


```{r}
alpha_coral$Region<-as.factor(alpha_coral$Region) #make sure region is a factor 

kruskal.test(rich1~Region,data=alpha_coral) #run a kruskal wallis test of richness by region 

```
Ok, so it's significant, but how do you now test for differences between specific regions like we were able to do with the Tukey? You use multiple kruskal wallis test and adjust the pvalues for multiple comparisons.

Let's run through an example with richness:
```{r}
#set it up just like we set up the multiple permanova test
sampledf<-data.frame(sample_data(ps))
sampledf$Region<-as.factor(sampledf$Region)

region.list<-levels(sampledf$Region)
region.list

pvals<-c()
locations<-c()
for (i in 1:length(region.list)){
  region.leftout<-region.list[i]
  alph_sub<-subset(alpha_coral,Region!=region.leftout) #subset the alpha metrics to just two locations
  kwt<-kruskal.test(rich1~Region,alph_sub) #kruskal wallis test
  print(kwt) 
  pvals<-c(pvals,kwt$'p.value') #save the pval 
  locations<-rbind(locations,c(region.list[-i]))#save the locations tested
}
combinedtest<-cbind(locations,pvals)
mult_kwt<-data.frame(combinedtest)
mult_kwt$pvaladj<-p.adjust(mult_kwt$pvals,method="bonferroni")
mult_kwt
```





