---
title: "Relative Abundance and Bubble plots"
output: html_notebook
---

# Relative Abundance and Bubble Plots

### Relative abundance plots vs bubble plots:
Both of these plots show the relative abundance of at least some taxa. Beyond the obvious difference of one being a stacked barplot and the other being a bubble plot, the key difference between the two is that relative abundance plots show the whole story of what is present in the samples, whereas bubble plots only show the key taxa, usually the significant taxa pulled out by a simper, ancom, or corncob analysis. Relative abundance plots are most useful when you want to look at compositional differences among samples, they can also give a very rough sense of the diversity in your samples. Bubble plots are most useful when you want to look at differences in abundance of key taxa. However, the bar plots can also pick out key taxa and just visualize those.

## load packages
```{r}
library("ggplot2")
library("phyloseq")
library("RColorBrewer")
library("reshape2")
library("nlme")
library("dplyr")
library("compositions")
library("tidyverse")
```

## load the files needed
```{r}
ps <- readRDS("examplePhyloseq.rds") #import the .rds phyloseq object made in the intro to phyloseq notebook
ps #confirm it worked

```

### Add information ot taxonomy
```{r}
tax.clean<-data.frame(tax_table(ps))
### Ok, so some ASVs don't have classification at all levels, we can use the following for loop to add some info
for (i in 1:nrow(tax.clean)){
  if (tax.clean[i,2] == ""){
    kingdom <- paste("Kingdom_", tax.clean[i,1], sep = "")
    tax.clean[i, 2:7] <- kingdom
    } else if (tax.clean[i,3] == ""){
      phylum <- paste("Phylum_", tax.clean[i,2], sep = "")
      tax.clean[i, 3:7] <- phylum
    } else if (tax.clean[i,4] == ""){
      class <- paste("Class_", tax.clean[i,3], sep = "")
      tax.clean[i, 4:7] <- class
    } else if (tax.clean[i,5] == ""){
      order <- paste("Order_", tax.clean[i,4], sep = "")
      tax.clean[i, 5:7] <- order
    } else if (tax.clean[i,6] == ""){
      family <- paste("Family_", tax.clean[i,5], sep = "")
      tax.clean[i, 6:7] <- family
    } else if (tax.clean[i,7] == ""){
      tax.clean$Species[i] <- paste("Genus",tax.clean$Genus[i], sep = "_")
    }
}
head(tax.clean)
taxonomy2<-as.matrix(tax.clean)
# add taxonomy to ps
tax_table(ps)<-taxonomy2

```

## Relative Abundance plots

What goes in: a phyloseq object
What goes out: a relative abundance plot

### Get the data in the right format for plotting
```{r}
#merge taxa on the class level, so that there's only 1 value for each Class CHANGE HERE IF YOU WANT A DIFFERENT TAXONOMIC LEVEL
physeq.f.glom <- tax_glom(ps, "Class",NArm=FALSE) 

#transform data from raw abundance to relative abundance
physeq.f.ra.glom <- transform_sample_counts(physeq.f.glom, function(x) x*100/sum(x)) 

#make otu table a dataframe
propData <- as.data.frame((otu_table(physeq.f.ra.glom))) 
head(propData)

#set up your dataframe that will be used to make the relative abundance plot
plotDF1 <- propData %>% #within the propData variable
  rownames_to_column(var="taxa") %>% #set the rownames (sample IDS) to a column
  melt() %>% #takes the data from wide to long format
  magrittr::set_names(c("sample", "taxa", "relab")) #rename columns
head(plotDF1)#inspect to make sure it worked

#extract the taxonomy info into something that can be combined with the plot dataframe
taxonomy.2 <- tax_table(physeq.f) %>%
  as.data.frame() %>%
  rownames_to_column(var="taxa") #add the rownames to a column called taxa
head(taxonomy.2)#just checking it worked

#combine dataframes merging by taxa, with the columns listed in select. ALSO CHANGE CLASS HERE IF YOU WANT A DIFFERENT TAXONOMIC RANK
plotDF1 <- left_join(plotDF1, taxonomy.2, by="taxa") %>% select(taxa, sample, relab, Class) 
head(plotDF1)#just checking it worked

sampdata<-data.frame(sample_data(physeq.f)) #make dataframe of the sample data
metadata.2 <- sampdata %>%
  rownames_to_column(var="sample") #make a column of rownames (sample)
head(metadata.2)

#combine plotDF1 and metadata.2 by sample name
plotDF1 <- left_join(plotDF1, metadata.2, by="sample") 
head(plotDF1) #just checking it worked

#check class of things that should be factors. ALSO CHANGE CLASS HERE IF YOU WANT A DIFFERENT TAXONOMIC RANK
plotDF1$sample<-factor(plotDF1$sample) 
plotDF1$Site<-factor(plotDF1$Region) 

#ASVs were first glommed and then counts were transformed into relative abundaces. ITS2 types that made up less than 5% of a sample's relative abundance were grouped into a <5% category to simplify the visualization in Figure 7. 
#get rid of those <5% 
plotDF1[plotDF1$relab <1, ]$Class<-"<1% abundance"
head(plotDF1)
plotDF1$Class<-factor(plotDF1$Class) 


```

### Setup your colors
```{r}
#check how many classes there are to determine how many colors you need
length(levels(factor(plotDF1$Class))) 

#make a list of colors the same length as the number of classes
mycolors <- c("grey","red","orange","yellow","green","blue","magenta","darkorchid4") 

#set the colors to an associated taxonomic rank. ALSO CHANGE CLASS HERE IF YOU WANT A DIFFERENT TAXONOMIC RANK
names(mycolors)<-levels(factor(plotDF1$Class)) 
mycolors #confirm it worked

```

### Plot!
```{r}
#make a plot with genotype/sample name on the x-axis, relative abundance on the y-axis, and color filled by Class. ALSO CHANGE CLASS HERE IF YOU WANT A DIFFERENT TAXONOMIC RANK
ggplot(plotDF1, aes(x=plotDF1$sample, relab, fill=Class)) + 
  #make it a barplot with classes stacked on top of each other
  geom_bar(stat="identity", position = "stack") + 
  #set x-axis labels to be oriented vertically, and set the size of the text to 10
  theme(axis.text.x = element_text(angle = 90,size = 6.5)) + 
  #set the colors being filled in by class to the mycolors variable made earlier
  scale_fill_manual(values = mycolors,limits = levels(plotDF1$Class)) + 
  #facet wrap (separate out) by Region
  facet_grid(~Region,space="free_x",scales="free_x",drop=TRUE) + 
  #set legend title size to 8 and legend text size to 6
  theme(legend.title = element_text(size = 8),legend.text = element_text(size = 6),legend.position = "bottom")+
  xlab("")+
  ylab("% Relative Abundance")

```

## Bubble plots
Bubble plots with ASVs on the y-axis

### Bubble plot with ASVs on the y-axis

What goes in: a phyloseq object
What goes out: a bubble plot where significant ASVs is the y-axis
ASVs pulled from simper
```{r}
#switch to relative abundance from raw abundance
physeq.f.ra <- transform_sample_counts(ps, function(x) x*100/sum(x)) 

#merge by region
p<-merge_samples(physeq.f.ra,"Region")

o<-otu_table(p) #make variable of p's otu table

rowSums(otu_table(physeq.f.ra)) #confirm that all of the rows add up to 100 (this makes sure that the data is actually in relative abundance format)

propData <- as.data.frame(o) #make otu table a dataframe

t<-data.frame(cbind(levels(as.factor(sample_data(ps)$Region)))) #make dataframe with region

t <-cbind(t,propData$ASV_2,propData$ASV_3,propData$ASV_4) #add the abundance data of the 3 ASVs from the simper to the dataframe 

colnames(t)<-c("Region","ASV_2","ASV_3","ASV_4") #set the column names

pcm <- melt(t, id = c("Region")) #convert data frame from a "wide" format to a "long" format

#make plot with Region as the x-axis, the significant ASVs as the y-axis, and color filled by the significant ASVs
ggplot(pcm, aes(x = Region, y = variable,fill=variable)) + 
  #make bubble plot format where the size of the bubble is the relative abundance
  geom_point(aes(size = as.numeric(value)), alpha = 0.75, shape = 21) +  
  #set bubble size scaling
  scale_size_continuous(limits = c(0.000001, 100), range = c(.5,10), breaks = c(1,25,50,75)) + 
   #set labels for size legend and color legend
  labs( x= "", y = "", size = "Abundance", fill = "Significant ASVs")  +
  #set various text sizes and colors, as well as legend position
  theme(legend.key=element_blank(), axis.text.x = element_text(colour = "black", size = 12, face = "bold", angle = 90, vjust = 0.3, hjust = 1), axis.text.y = element_text(colour = "black", face = "bold", size = 11), legend.text = element_text(size = 10, face ="bold", colour ="black"), legend.title = element_text(size = 12, face = "bold"), panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2), legend.position = "right") + 
  #set y-scale limits to significant ASVs
  scale_y_discrete(limits = rev(levels(pcm$variable))) 

taxdf<-data.frame(tax_table(physeq.f.ra))
asvnames<-taxdf[rownames(taxdf) %in% c("ASV_2","ASV_3","ASV_4"), ]$Species

ggplot(pcm, aes(x = Region, y = variable,fill=variable)) + 
  #make bubble plot format where the size of the bubble is the relative abundance
  geom_point(aes(size = as.numeric(value)), alpha = 0.75, shape = 21) +  
  #set bubble size scaling
  scale_size_continuous(limits = c(0.000001, 100), range = c(.5,10), breaks = c(1,25,50,75)) + 
   #set labels for size legend and color legend
  labs( x= "", y = "", size = "Abundance", fill = "Significant ASVs")  +
  #set various text sizes and colors, as well as legend position
  theme(legend.key=element_blank(), axis.text.x = element_text(colour = "black", size = 12, face = "bold", angle = 90, vjust = 0.3, hjust = 1), axis.text.y = element_text(colour = "black", face = "bold", size = 11), legend.text = element_text(size = 10, face ="bold", colour ="black"), legend.title = element_text(size = 12, face = "bold"), panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2), legend.position = "right") + 
  #set y-scale limits to significant ASVs
  scale_y_discrete(limits = rev(levels(pcm$variable)),labels=asvnames) #add labels here

```

